---
title: "Case_Study1_cleaned"
output:
  pdf_document: default
  html_document: default
date: "2025-01-12"
---

###Introduction
Welcome to the Cyclistic bike-share analysis case study! In this case study, you work for a
fictional company, Cyclistic, along with some key team members. In order to answer the
business questions, follow the steps of the data analysis process: Ask, Prepare, Process,
Analyze, Share, and Act. Along the way, the Case Study Roadmap tables — including guiding
questions and key tasks — will help you stay on the right path.
###Scenario
You are a junior data analyst working on the marketing analyst team at Cyclistic, a bike-share
company in Chicago. The director of marketing believes the company’s future success
depends on maximizing the number of annual memberships. Therefore, your team wants to
understand how casual riders and annual members use Cyclistic bikes differently. From these
insights, your team will design a new marketing strategy to convert casual riders into annual
members. But first, Cyclistic executives must approve your recommendations, so they must be
backed up with compelling data insights and professional data visualizations.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Analyze**

### **1. Combine Columns and Data**

```{r combine-columns}
# Set working directory
setwd("/Users/giuliaribeiro/Documents/R_course/Case_Study1/")

# Import monthly datasets
library(dplyr)
library(readr)
library(lubridate)

data_dir <- "./monthly_files/"
file_list <- list.files(path = data_dir, pattern = "*.csv", full.names = TRUE)

# Combine datasets
data_combined <- file_list %>%
  lapply(read_csv) %>%
  bind_rows() %>%
  mutate(month = month(started_at),
         month_name = month(started_at, label = TRUE, abbr = FALSE))

# Preview combined data
glimpse(data_combined)
```
### **2. Clean and Transform Data**

I calculate the percentage of missing data on the columns to understand the strategy to be adopted. I saw that for station names 18% was missing data, so in that case to not miss a lot of data I just replaced NA for Unknown. In the case of end_lat and end_long it was only 0.1%, so In this case I dropped the rows with NA.


```{r}
# Calculate the total number of rows in the dataset
total_rows <- nrow(data_combined)

# Calculate the number of missing values for each column
missing_values <- colSums(is.na(data_combined))

# Calculate the percentage of missing values for each column
missing_percentage <- (missing_values / total_rows) * 100

# Combine the results into a data frame for better readability
missing_summary <- data.frame(
  Column = names(missing_values),
  Missing_Count = missing_values,
  Missing_Percentage = round(missing_percentage, 1) # Rounded to one decimal place
)

# Print the summary
print(missing_summary)

```

```{r clean-transform}
# Handle missing values
library(tidyr)
data_cleaned <- data_combined %>%
  mutate(
    start_station_name = replace_na(start_station_name, "Unknown"),
    start_station_id = replace_na(start_station_id, "Unknown"),
    end_station_name = replace_na(end_station_name, "Unknown"),
    end_station_id = replace_na(end_station_id, "Unknown"),
    ride_length = as.numeric(difftime(ended_at, started_at, units = "mins")),
    day_of_week = wday(started_at, label = TRUE)
  ) %>%
  drop_na(end_lat, end_lng) # Drop rows with critical missing values

# Preview cleaned data
glimpse(data_cleaned)
```

I added two new columns that I think will give usefull insights in this analysis:
ride_length and day_of_week

### **3. Descriptive Analysis**

```{r}
# Analyze ride length (e.g., by month)
monthly_summary <- data_cleaned %>%
  group_by(month, member_casual) %>%
  summarize(
    avg_ride_length = mean(ride_length, na.rm = TRUE),
    total_rides = n(),
    .groups = "drop"
  )

# Print summary
print(monthly_summary)

# Visualize ride length over the months
library(ggplot2)

ggplot(monthly_summary, aes(x = month, y = avg_ride_length, color = member_casual)) +
  geom_line() +
  geom_point() +
  labs(title = "Average Ride Length Over Months", x = "Month", y = "Average Ride Length (mins)") +
  scale_x_continuous(breaks = 1:12, labels = month.name) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r descriptive-analysis}
# Calculate mean and max ride length
mean_ride_length <- mean(data_cleaned$ride_length, na.rm = TRUE)
max_ride_length <- max(data_cleaned$ride_length, na.rm = TRUE)

# Convert day_of_week to character (remove ordering)
data_cleaned <- data_cleaned %>%
  mutate(day_of_week = as.character(day_of_week))

# Calculate mode of day_of_week
day_mode <- data_cleaned %>%
  group_by(day_of_week) %>%
  summarize(count = n(), .groups = "drop") %>%  # Ensure proper grouping behavior
  arrange(desc(count)) %>%
  slice_head(n = 1) %>%  # Select the first row explicitly
  pull(day_of_week)      # Extract the mode value

print(day_mode)

list(mean_ride_length = mean_ride_length, max_ride_length = max_ride_length, day_mode = day_mode)

```
### **4. Pivot Table Analysis**

```{r pivot-analysis}
# Average ride length by member type
data_summary <- data_cleaned %>%
  group_by(member_casual) %>%
  summarize(avg_ride_length = mean(ride_length, na.rm = TRUE))

# Average ride length by day of week
data_by_day <- data_cleaned %>%
  group_by(day_of_week, member_casual) %>%
  summarize(avg_ride_length = mean(ride_length, na.rm = TRUE))

# Count rides by day of week
ride_count <- data_cleaned %>%
  group_by(day_of_week, member_casual) %>%
  summarize(total_rides = n())

list(summary = data_summary, by_day = data_by_day, ride_count = ride_count)
```

### **5. Seasonal Analysis**

```{r seasonal-analysis}
# Filter for summer months (June, July, August)
summer_data <- data_cleaned %>%
  filter(month(started_at) %in% c(6, 7, 8))

# Descriptive stats for summer
summer_stats <- summer_data %>%
  group_by(member_casual) %>%
  summarize(
    avg_ride_length = mean(ride_length, na.rm = TRUE),
    total_rides = n()
  )

summer_stats
```
### **6. Visualization**

```{r visualizations, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)

# Plot 1: Average Ride Length by Member Type
ggplot(data_summary, aes(x = member_casual, y = avg_ride_length, fill = member_casual)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Ride Length by Member Type", x = "User Type", y = "Average Ride Length (mins)") +
  theme_minimal()

# Plot 2: Total Rides by Day of Week and Member Type
ggplot(ride_count, aes(x = day_of_week, y = total_rides, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Total Rides by Day of Week and Member Type", x = "Day of Week", y = "Total Rides") +
  theme_minimal()

# Plot 3: Average Ride Length by Day of Week and Member Type
ggplot(data_by_day, aes(x = day_of_week, y = avg_ride_length, color = member_casual, group = member_casual)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Average Ride Length by Day of Week and Member Type", x = "Day of Week", y = "Average Ride Length (mins)") +
  theme_minimal()
```

### **7. Export Summary File**

```{r export-summary}
# Export cleaned and summarized data
write_csv(data_cleaned, "cleaned_cyclistic_data.csv")
write_csv(data_summary, "ride_length_summary.csv")
write_csv(data_by_day, "ride_length_by_day.csv")
```


### **8. Summary Narrative**

#### Insights from Analysis

1. **Key Trends:**
   - Casual riders consistently have longer average ride lengths compared to members, particularly during weekends and summer months.
   - Members primarily use bikes during weekdays, suggesting frequent utility-focused usage such as commuting.

2. **Seasonal Variations:**
   - Summer months show a significant spike in ridership, especially among casual users, indicating a strong link to leisure activities.

3. **Most Popular Days:**
   - Saturdays dominate as the most popular day for riding, particularly for casual users.

#### Implications for Marketing Strategy

- **Targeting Casual Riders:** Highlight benefits of membership during summer and weekends to encourage conversions. For example, offer discounts for summer memberships.
- **Member Retention:** Reinforce the utility and cost-effectiveness of membership for regular commuters.
- **Seasonal Campaigns:** Plan promotions during peak summer months to attract more casual users.

